Analyzing the Pakistan Cricket Team's Performance (2002-2024)

### 2024-12-04 paper/paper.qmd
Added results section to paper

---
title: "Analyzing the Pakistan Cricket Team's Performance (2002-2024)"
subtitle: "My subtitle if needed"
author: 
  - Muhammad Abdullah Motasim
thanks: "Code and data are available at: [https://github.com/abdullah-motasim/Analyzing-Pakistans-Cricket-Data](https://github.com/abdullah-motasim/Analyzing-Pakistans-Cricket-Data)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
table-of-contents: true
toc-depth: 2
---

@@

library(tidyverse)
library(here)
library(ggplot2)
library(arrow)
library(patchwork)

pakistan_match_data <- read_parquet(here("data/02-analysis_data/cleaned_pakistan_match_data.parquet"))
```

# Introduction
Pakistan’s cricket team holds a storied legacy, marked by remarkable highs and challenging lows. From their 1992 World Cup triumph to fluctuations in form over the decades, analyzing trends in performance offers insights into their journey and areas for strategic improvement.

@@

## Estimand
The estimand is the probability that Pakistan wins a cricket match, given the opponent, toss outcome, and year.

@@

The data for this study were obtained from the cricketdata R package [@cricketdata] and all analysis was performed using `R` [@citeR] alongside the following packages: `tibble` [@tibble], `readr` [@readr], `arrow` [@arrow], `tidyverse` [@tidyverse], `dplyr` [@dplyr], `testthat` [@testthat], `caret` [@caret], `rstanarm` [@rstanarm], `tidyr` [@tidyr], patchwork [@patchwork], and ggplot2 [@ggplot2].

@@

## Raw Data
The cricketdata library offers a variety of functions that provide data sets on different cricket statistics. However, for our analysis, we focus on two key areas: individual match statistics, to determine which matches Pakistan played in and how they performed, and player career statistics, to assess how the quality of players has evolved over time and its impact on Pakistan's overall performance. To extract the relevant data, we primarily used two functions from the library: fetch_cricinfo, which retrieves individual career performance data, and fetch_cricsheet, which provides detailed match information.

## Data Cleaning
The raw match data initially contained 820 observations across 25 variables, each detailing a match between two countries. The data cleaning process involved selecting only matches in which Pakistan played, and retaining only the variables relevant to our analysis. These variables include:

@@

## Variable Analysis
Figures @fig-match-stats-1 and @fig-match-stats-2 display the count of several important match statistics. From these, we can observe that Pakistan faces a variety of opponents, with the most frequent ones being England, Sri Lanka, and Australia. Notably, Pakistan has had a challenging record against Australia, losing a significant number of matches, while many of their matches ended in draws.

@@

```{r}
#| label: fig-match-stats-1
#| fig-cap: Counts for match variables of interst
#| echo: false

@@

# Figure 2: Distribution of matches played by Team2
fig2 <- ggplot(data, aes(x = team2)) +
  geom_bar(fill = "coral") +
  labs(title = "Pakistan Opponent Team", x = "Team", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Figure 3: Match outcomes (Winner)
fig3 <- ggplot(data, aes(x = winner)) +
  geom_bar(fill = "darkgreen") +
  labs(title = "Matches Won by Each Team", x = "Team", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Figure 4: Distribution of Toss Winners
fig4 <- ggplot(data, aes(x = toss_winner)) +
  geom_bar(fill = "purple") +
  labs(title = "Tosses Won by each Team", x = "Team", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Figure 5: Pakistan Win Outcomes
fig5 <- ggplot(data, aes(x = pakistan_win)) +
  geom_bar(fill = "gold") +
  labs(title = "Matches Won by Pakistan", x = "Outcome", y = "Count") +
  scale_x_discrete(labels = c("0" = "Lost", "1" = "Won")) +  # Change 1 and 0 to Won and Lost
  theme_minimal()

# Figure 6: Toss Win Outcomes
fig6 <- ggplot(data, aes(x = toss_win)) +
  geom_bar(fill = "dodgerblue") +
  labs(title = "Tosses Won by Pakistan", x = "Outcome", y = "Count") +
  scale_x_discrete(labels = c("0" = "Lost", "1" = "Won")) +  # Change 1 and 0 to Won and Lost
  theme_minimal()

# Figure 7: Match Results by Year
fig7 <- ggplot(data, aes(x = year)) +
  geom_bar(fill = "brown") +
  labs(title = "Pakistan's Matches Played \nper Year", x = "Year", y = "Number of Matches") +
  theme_minimal()

combines_plot1 <- (fig2)/(fig3)
print(combines_plot1)

@@

```{r}
#| label: fig-match-stats-2
#| fig-cap: Counts for match variables of interst
#| echo: false

combines_plot2 <- (fig4 | fig5) /( fig6 | fig7)
print(combines_plot2)
```

@@

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

@@

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

# library(rstanarm)
# 
# first_model <-
#   readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false
# 
# modelsummary::modelsummary(
#   list(
#     "First model" = first_model
#   ),
#   statistic = "mad",
#   fmt = 2
# )
```

@@

# Appendix {-}

@@

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
# 
# pp_check(first_model) +
#   theme_classic() +
#   theme(legend.position = "bottom")
# 
# posterior_vs_prior(first_model) +
#   theme_minimal() +
#   scale_color_brewer(palette = "Set1") +
#   theme(legend.position = "bottom") +
#   coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

# plot(first_model, "trace")
# 
# plot(first_model, "rhat")
```

>>>>

---
title: "Analyzing the Pakistan Cricket Team's Performance (2002-2024)"
subtitle: "My subtitle if needed"
author: 
  - Muhammad Abdullah Motasim
thanks: "Code and data are available at: [https://github.com/abdullah-motasim/Analyzing-Pakistans-Cricket-Data](https://github.com/abdullah-motasim/Analyzing-Pakistans-Cricket-Data)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
table-of-contents: true
toc-depth: 2
editor: 
  markdown: 
    wrap: 72
---

@@

library(tidyverse)
library(here)
library(ggplot2)
library(arrow)
library(modelsummary)
library(tinytable)
library(knitr)
library(kableExtra)

player_data <- read_parquet("data/02-analysis_data/cleaned_player_data.parquet")

pakistan_match_data <- read_parquet(here("data/02-analysis_data/cleaned_pakistan_match_data.parquet"))
model <- readRDS(here("models/win_chance.rds"))
confusion_matrix <- readRDS(here("models/confusion_matrix_win_chance.rds"))

# Introduction

Pakistan’s cricket team holds a storied legacy, marked by remarkable
highs and challenging lows. From their 1992 World Cup triumph to
fluctuations in form over the decades, analyzing trends in performance
offers insights into their journey and areas for strategic improvement.

@@

## Estimand

The estimand is the probability that Pakistan wins a cricket match,
given the opponent, toss outcome, and year.

@@

The data for this study were obtained from the cricketdata R package
[@cricketdata] and all analysis was performed using `R` [@citeR]
alongside the following packages: `tibble` [@tibble], `readr` [@readr],
`arrow` [@arrow], `tidyverse` [@tidyverse], `dplyr` [@dplyr], `testthat`
[@testthat], `caret` [@caret], `rstanarm` [@rstanarm], `tidyr` [@tidyr],
modelsummary [@modelsummary], tinytable [@tinytable], knitr [@knitr], and ggplot2 [@ggplot2].

@@

## Raw Data

The cricketdata library offers a variety of functions that provide data
sets on different cricket statistics. However, for our analysis, we
focus on two key areas: individual match statistics, to determine which
matches Pakistan played in and how they performed, and player career
statistics, to assess how the quality of players has evolved over time
and its impact on Pakistan's overall performance. To extract the
relevant data, we primarily used two functions from the library:
fetch_cricinfo, which retrieves individual career performance data, and
fetch_cricsheet, which provides detailed match information.

## Data Cleaning

The raw match data initially contained 820 observations across 25
variables, each detailing a match between two countries. The data
cleaning process involved selecting only matches in which Pakistan
played, and retaining only the variables relevant to our analysis. These
variables include:

@@

## Variable Analysis

Figures @fig-match-stats displays the count of several important match
statistics. From these, we can observe that Pakistan faces a variety of
opponents, with the most frequent ones being England, Sri Lanka, and
Australia. Notably, Pakistan has had a challenging record against
Australia, losing a significant number of matches, while many of their
matches ended in draws.

@@

```{r}
#| label: fig-match-stats
#| fig-cap: Counts for match variables of interst
#| fig-subcap: ["Pakistan Opponent Team","Matches Won by Each Team",
#| "Tosses Won by each Team", "Matches Won by Pakistan", "Tosses Won by Pakistan", "Pakistan's Matches Played per Year"]
#| layout-ncol: 2
#| layout-nrow: 3
#| echo: false

@@

# Figure 1: Distribution of matches played by Team2
ggplot(data, aes(x = team2)) +
  geom_bar(fill = "coral") +
  labs(x = "Team", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Figure 2: Match outcomes (Winner)
ggplot(data, aes(x = winner)) +
  geom_bar(fill = "darkgreen") +
  labs(x = "Team", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Figure 3: Distribution of Toss Winners
ggplot(data, aes(x = toss_winner)) +
  geom_bar(fill = "purple") +
  labs(x = "Team", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Figure 4: Pakistan Win Outcomes
ggplot(data, aes(x = pakistan_win)) +
  geom_bar(fill = "gold") +
  labs(x = "Outcome", y = "Count") +
  scale_x_discrete(labels = c("0" = "Lost", "1" = "Won")) +  # Change 1 and 0 to Won and Lost
  theme_minimal()

# Figure 5: Toss Win Outcomes
ggplot(data, aes(x = toss_win)) +
  geom_bar(fill = "dodgerblue") +
  labs(x = "Outcome", y = "Count") +
  scale_x_discrete(labels = c("0" = "Lost", "1" = "Won")) +  # Change 1 and 0 to Won and Lost
  theme_minimal()

# Figure 6: Match Results by Year
ggplot(data, aes(x = year)) +
  geom_bar(fill = "brown") +
  labs(x = "Year", y = "Number of Matches") +
  theme_minimal()

@@

@@

I used a Bayesian Logistic Regression model to estimate the probability
of Pakistan winning a cricket match based on several key factors.
Logistic regression is a statistical technique used for binary
classification, where the goal is to predict the probability of a
categorical outcome. In this case, the outcome is whether Pakistan wins
or loses a match.

## Model Overview

The model incorporates three independent variables that may influence
match outcomes: year, opponent, and toss winner. The dependent variable
is match_outcome, where 1 indicates a win for Pakistan and 0 indicates a
loss. This setup allows us to examine the influence of key variables on
Pakistan's performance.

By incorporating these variables, the model explores how factors such as
the year of the match, the opposing team, and whether Pakistan won the
toss (which influences the decision to bat or bowl) affect Pakistan’s
likelihood of winning. This approach helps uncover trends in Pakistan's
performance over time and under varying conditions, providing insights
into factors that may inform future match strategies.

## Model Equation

The model I will be using can be described using the following equation:

```{=tex}
\begin{equation}
\log\left(\frac{\hat{p}}{1 - \hat{p}}\right) = \beta_0 + \beta_1 \times \text{year} + \beta_2 \times \text{opponent} + \beta_3 \times \text{toss winner} 
\end{equation}
```
$$
\begin{aligned}
\beta_0 & \sim \mbox{Normal}(0, 2.5)\\
\beta_1 & \sim \mbox{Normal}(0, 2.5)\\
\beta_2 & \sim \mbox{Normal}(0, 2.5)\\
\beta_3 & \sim \mbox{Normal}(0, 2.5)
\end{aligned}
$$ where,

-   $\hat{p}$ is the predicted probability of Pakistan winning a match.
-   $\beta_0$ is the intercept (the log-odds of winning when all
    predictors are at their reference level).
-   $\beta_1$ is the coefficient for the year variable, representing how
    match outcomes are affected by the year of the match.
-   $\beta_2$ are the coefficients for each level of the categorical
    variable team2 (the opponent), with the reference category excluded.
    This allows the model to compare the odds of winning against
    different opponents.
-   $\beta_3$ is the coefficient for the toss_win variable (whether
    Pakistan won the toss).

## Model Priors

The priors used in the model represent our assumptions about the
parameters before observing the data. These priors are chosen based on
reasonable beliefs about the likely values of the coefficients and are
specified as follows:

-   `Prior for the Coefficients (year, team2, toss_win)`: Each of the
    model coefficients is assigned a Normal prior with a mean of 0 and a
    standard deviation of 2.5. This reflects the assumption that we
    expect the effects of the predictors to be centered around zero,
    with a reasonable amount of uncertainty (a wide spread due to the
    scale of 2.5) about their true values.

-   `Prior for the Intercept`: A normal prior with the same parameters
    as the coefficients (mean = 0, scale = 2.5) is used for the
    intercept. The intercept represents the baseline log-odds of
    Pakistan winning a match when all other predictors are zero.

-   `Prior for the Auxiliary Parameter (prior_aux)`: This parameter
    represents the dispersion of the logistic regression model. An
    Exponential prior with a rate of 1 is used, which is typical in
    Bayesian models for logistic regression. The Exponential prior
    allows for a distribution with only non-negative values, reflecting
    uncertainty in the variance of match outcomes, with higher values
    suggesting more variability in outcomes.

The Bayesian approach incorporates prior knowledge (through these
priors) and updates it with the data from the training set, enabling the
model to estimate relationships while accounting for uncertainty in the
coefficients.

## Model Justification

A decision tree model was also considered, which could capture
non-linear relationships and interactions between predictors. However,
decision trees are more prone to overfitting and can be less
interpretable than logistic regression, especially in cases with
categorical variables like team2.

An alternative approach considered was a frequentist logistic regression
model. This model would not incorporate prior information and would
instead focus purely on the likelihood of the data. However, the
Bayesian approach was chosen due to its ability to incorporate prior
knowledge and handle uncertainty more effectively, especially in the
context of small sample sizes and the complexity of cricket data.

This model provides a useful framework for estimating Pakistan's
likelihood of winning a match based on key factors. By using a Bayesian
Logistic Regression approach, the model incorporates uncertainty in the
estimates, offers a clear interpretation of the effects of each
predictor, and is flexible enough to adapt to new data as it becomes
available. However, it should be acknowledged that there are limitations
regarding omitted variables and potential model assumptions that could
be addressed in future iterations.

This is an overview of the model. More details such as assumptions,
limitations, convergence, and validation can be found in
@sec-model-details

@@

Our results are summarized in @tbl-model-results, and @fig-model-results
presents the predicted values of the coefficients. This figure also
includes the 95% confidence intervals for the predictors, as estimated
by our Bayesian model. A 95% confidence interval indicates that there is
a 95% probability that the true value of the parameter lies within the
specified range, given the observed data. In the graph, each point
represents the predicted value, and the horizontal line extending from
each point shows the 95% confidence interval.

Variables to the right of 0 have a positive correlation with the
outcome, meaning that increasing these values increases the probability
of Pakistan winning. Conversely, variables to the left of 0 exhibit a
negative correlation, meaning that increasing these variables reduces
the probability of Pakistan winning a match.

Finally, it is important to note that @fig-model-results is plotted
without the intercept term and the team2Ireland term. As shown in
@tbl-model-results, these two values have much larger magnitudes
compared to the other predictors. If included in the plot, their scale
would be so large that the smaller variations in the other coefficients
would be obscured.

```{r}
#| label: tbl-model-results
#| tbl-cap: "Coefficient estimates and standard errors"
#| echo: false
#| eval: true
#| warning: false

 modelsummary::modelsummary(
   list("Model" = model),
   statistic = "mad",
   fmt = 2,
   gof_omit = "Num.Obs.|R2|Log.Lik.|RMSE|ELPD|LOOIC|LOOIC s.e.|WAIC"
 )
```

```{r}
#| echo: false
#| eval: true
#| label: fig-model-results
#| fig-cap: "Coefficient estimates and 95% confidence intervals"
#| warning: false

 modelplot(model, coef_omit = "Intercept|team2Ireland") + 
   labs(x = "Coefficient Estimate", y="Predictor Variable")
```

Analyzing some of these coefficients further, The results from the
regression model are as follows:

-   **Intercept**: The intercept of -55.94 (with a standard error of
    80.63) represents the log-odds of Pakistan winning a match when all
    other variables are set to their reference categories. This value is
    quite large in magnitude, but it is not directly interpretable in
    isolation. The negative value suggests a very low baseline
    probability of winning in the absence of other influencing factors.

-   **Year**: The coefficient for year is 0.03 (standard error = 0.04).
    This positive coefficient suggests that, as time progresses, the
    probability of Pakistan winning a match increases very slightly.
    However, the small magnitude of the coefficient indicates that this
    effect is minimal. The relatively narrow standard error suggests
    that the estimate is reasonably precise, but the practical
    significance of this trend is modest.

-   **Opponent**: The coefficients for the opponent variable represent
    how the likelihood of Pakistan winning a match changes when facing
    different teams. These coefficients should be interpreted relative
    to the reference team, which is not explicitly listed here but is
    assumed to be the team with the lowest or most neutral effect:

    -   `team2Bangladesh` has a coefficient of 3.62 (standard error =
        1.31), which indicates a significant positive association with
        Pakistan’s success. Matches against Bangladesh appear to
        increase Pakistan's chances of winning.

    -   `team2Sri Lanka` shows a positive coefficient of 1.09 (standard
        error = 0.72), suggesting that matches against Sri Lanka are
        associated with a higher probability of Pakistan winning
        compared to the reference team, though the effect is less
        pronounced than for Bangladesh.

    -   The coefficients for other teams, such as `team2India` (0.24, standard error = 1.44) and `team2South Africa` (0.29, standard error = 0.90), are much smaller, indicating weaker associations with match success.
    - `team2Ireland` has the largest coefficient (19.77, standard error = 15.67), indicating a very high probability of success when playing against Ireland. However, this coefficient also has a large standard error, meaning there is significant uncertainty around this estimate.

- `Toss Win`: The coefficient for toss_win is 0.60 (standard error = 0.45), suggesting that winning the toss slightly increases Pakistan's likelihood of winning the match. This result supports the common understanding that the team winning the toss often gains an advantage in deciding whether to bat or bowl first.

```{r}
#| label: fig-opponent-stats
#| fig-cap: "Pakistan's wins and losses against all teams"
#| fig-subcap: ["Sri Lanka","New Zealand", "Australia", "England", "South Africa", "West Indies", "Zimbabwe", "Bangladesh", "Ireland", "India"]
#| layout-ncol: 2
#| layout-nrow: 5
#| echo: false
#| eval: true
#| warning: false

# Create a new column that accounts for wins, losses, and draws
pakistan_match_data <- pakistan_match_data %>%
  mutate(pakistan_result = case_when(
    winner == "Pakistan" ~ "Win",
    winner == "Draw" ~ "Draw",
    TRUE ~ "Loss"
  ))

# Loop over each opponent team
opponents <- unique(pakistan_match_data$team2)  # Get all unique opponents

# Create and save a plot for each opponent
for (opponent in opponents) {

  # Filter the data for the current opponent
  filtered_matches <- pakistan_match_data %>% filter(team2 == opponent)

  # Summarize the number of wins, losses, and draws
  win_loss_summary <- filtered_matches %>%
    group_by(pakistan_result) %>%
    summarise(count = n(), .groups = "drop")
  ggplot(data, aes(x = team2)) +
  geom_bar(fill = "coral") +
  labs(x = "Team", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  # Create the plot
  plot <- ggplot(win_loss_summary, aes(x = pakistan_result, y = count, fill = pakistan_result)) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(
      x = "Result",
      y = "Count",
      fill = "Result"
    ) +
    scale_fill_manual(values = c("Win" = "darkgreen", "Loss" = "brown2", "Draw" = "darkgoldenrod2")) +  # Custom colors for Win, Loss, Draw
    coord_cartesian(ylim = c(0, 14)) +
    theme_minimal()+
    theme(legend.position = "none")

print(plot)
}
```
@fig-opponent-stats illustrates the distribution of match outcomes (wins, losses, and draws) for Pakistan against eight different cricket teams: Sri Lanka, New Zealand, England, Australia, South Africa, West Indies, Zimbabwe, and Bangladesh. These results provide an overview of Pakistan's historical performance against each team and highlight patterns in their match outcomes.

- `Sri Lanka` (Figure 3a): Pakistan's matches against Sri Lanka are fairly balanced, with a slightly higher frequency of losses compared to wins. Draws are the least frequent outcome, suggesting a competitive dynamic between the two teams.

- `New Zealand` (Figure 3b): Matches against New Zealand reveal a similar trend, with losses being more frequent than wins. The proportion of draws remains relatively low, indicating fewer instances of matches being evenly contested.

- `England` (Figure 3c): Pakistan has a notably higher number of losses against England, with draws constituting the smallest category. This indicates that England has historically been a challenging opponent for Pakistan.

- `Australia` (Figure 3d): A significant disparity is observed in matches against Australia, where losses outnumber wins by a considerable margin. This trend underscores the dominance of Australia in these encounters.

- `South Africa` (Figure 3e): Matches against South Africa also show a higher proportion of losses compared to wins. However, the gap between these two outcomes is less pronounced compared to matches against Australia.

- `West Indies` (Figure 3f): Pakistan exhibits a stronger performance against West Indies, with wins slightly outnumbering losses. Draws remain the least frequent outcome, emphasizing decisive results in these encounters.

- `Zimbabwe` (Figure 3g): Against Zimbabwe, Pakistan demonstrates a clear dominance, with wins significantly outnumbering losses. Draws are nearly negligible, highlighting Pakistan's consistent success in these matches.

- `Bangladesh` (Figure 3h): Similar to Zimbabwe, Pakistan has a markedly higher number of wins against Bangladesh, with losses and draws being infrequent. This dominance underscores Pakistan's strong track record against this opponent.

Overall, these graphs provide valuable insights into Pakistan's historical performance trends across different opponents. While some teams, such as Australia and England, have posed significant challenges, others, like Bangladesh and Zimbabwe, have been more favorable matchups for Pakistan. These findings align with the regression model coefficients presented earlier, where the opponent variable was a significant predictor of Pakistan's success in matches. For instance, the high win rates against Bangladesh and Zimbabwe correspond to their positive regression coefficients, while the challenging outcomes against teams like Australia reflect negative or smaller coefficients.

```{r}
#| label: fig-pk-wins
#| fig-cap: "Number of Pakistan's Wins Over the Years"
#| echo: false
#| eval: true
#| warning: false

pakistan_win_data <- pakistan_match_data %>%
    mutate(pakistan_win = case_when(
    winner == "Pakistan" ~ 1,
    TRUE ~ 0))%>%
  mutate(year = lubridate::year(date))

# Summarize the number of wins per year
wins_by_year <- pakistan_win_data %>%
  group_by(year) %>%
  summarise(total_wins = sum(pakistan_win))

# Create a bar plot showing the number of wins per year
ggplot(wins_by_year, aes(x = year, y = total_wins)) +
  geom_bar(stat = "identity", fill = "darkblue", show.legend = FALSE) +  # Set a fixed color for all bars
  labs(
    x = "Year",
    y = "Total Wins"
  ) +
  theme_minimal()

```
@fig-pk-wins displays the number of wins Pakistan has achieved each year. From the graph, we observe that Pakistan had a modest performance from 2002 to 2010, with only 2-4 wins per year, followed by a sharp increase in 2011. After this surge, the team maintained an average of 4 wins per year, with some fluctuations. However, in 2019, there was a notable decline, with just one win recorded. Since this drop, Pakistan has struggled to reach their previous high of 4 wins per year, and the current average stands at approximately 2 wins per year.

@@

# Appendix {.unnumbered}

@@

## Model assumptions

-   `Independence of Observations`: We assume that the results of each
    match are independent. This assumption may be violated in cases
    where teams play multiple matches within a short period, but the
    model does not account for such temporal dependencies.

-   `Logistic Relationship`: The model assumes a linear relationship
    between the predictors (year, opponent, toss win) and the log-odds
    of Pakistan winning. This assumption may not capture more complex
    non-linear relationships, which would require more sophisticated
    models, such as generalized additive models (GAMs).

-   `No Interaction Effects`: The model currently does not include
    interaction terms (e.g., how the effect of winning the toss might
    differ by opponent). Interaction effects could be considered if they
    were hypothesized to be important, but they are omitted here for
    simplicity.

-   `Reference Category for team2`: The model assumes that the effect of
    playing against a reference opponent (the one excluded from the
    dummy variables for team2) is the baseline against which the other
    teams are compared. The choice of reference category is important
    and could influence the interpretation of the coefficients.

## Model limitations

-   `Omitted Variables`: The model does not account for other
    potentially important variables, such as player injuries, weather
    conditions, or venue specifics. These factors could influence the
    match outcome but are not included in the current model.

-   `Data Representation`: The model is based on matches involving
    Pakistan, so it does not generalize to other teams or international
    cricket broadly. Additionally, the dataset may have biases (e.g., a
    higher number of matches against certain opponents), which could
    affect the accuracy of the predictions.

-   `Data Sparsity`: The dataset available for analysis may not be large
    enough to fully capture all the complexities of cricket matches,
    especially with respect to less frequent events or rare outcomes.
    This could lead to uncertainty in the model estimates, particularly
    for smaller categories or outcomes that appear less frequently.

-   `Data Granularity`: The model uses year as a continuous variable.
    However, time trends in cricket might be better captured with more
    granular temporal variables (e.g., by series or tournament), as the
    year-by-year analysis may mask shorter-term trends.

## Model validation

We split the data into training and test sets to evaluate the predictive
performance of the models. The first model was trained on a subset of
the data, and its predictions were tested on the remaining matches in
the test set. Specifically, we used out-of-sample testing, where the
model was trained on the training set and evaluated on the test set to
assess its ability to generalize to new, unseen data.

For the first model, we predicted the probability of a match outcome
(e.g., whether Pakistan would win) for each match in the test set. Using
a threshold of 0.5, we classified these probabilities into binary
outcomes (1 for a predicted win, 0 for a predicted loss). To evaluate
model performance, we constructed a confusion matrix comparing the
predicted classes to the actual outcomes of the test set. From this
confusion matrix, we calculated accuracy as the proportion of correct
predictions (i.e., the sum of true positives and true negatives divided
by the total number of predictions). This provided an overall measure of
how well the model performed in predicting match outcomes. The result of
these tests is summarized in @tbl-model-metrics in the form of a
confusion matrix and accuracy.

We repeated the same procedure for the second model, predicting the
match outcome based on a different set of features (e.g., year and
opponent). The accuracy for both models was computed and reported
separately.

This approach provided a robust evaluation of each model's predictive
performance on unseen data, with the confusion matrix and accuracy
offering insights into the model’s classification capabilities. While
accuracy provides a good indication of overall model performance, we
also recognize that other metrics such as precision, recall, and F1
score might be more informative in the case of imbalanced classes.

```{r}
#| label: tbl-model-metrics
#| tbl-cap: Confusion matrix and Accuracy measurements for model
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| tbl-pos: 'H'
#| fig-align: center

# Print confusion matrix
print(confusion_matrix)

# Calculate accuracy
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)

# Print formatted accuracy with two decimal places
cat("\nModel Accuracy: ", sprintf("%.2f", accuracy), "\n", sep = "")
```

## Model diagnostics

**Fit Diagnostics**:

The mean posterior predictive distribution (mean_PPD) shown in @tbl-ppd is the mean of the posterior distribution of the predicted probabilities for Pakistan winning a match. The value of 0.4 indicates that, on average, the model predicts Pakistan has a 40% chance of winning a match based on the predictors.

**MCMC Diagnostics**:

@tbl-mcmc shows diagnostic measurements of the sampling process to assess
whether the model has converged and whether the samples are reliable.

-   `mcse (Monte Carlo Standard Error)`: The MCSE indicates the standard
    error of the Monte Carlo estimates for each parameter. Smaller
    values are better, as they show that the estimates are stable and
    not too variable. Most of the parameters here have small MCSE values
    (close to 0), indicating stable estimates.

-   `Rhat`: Rhat is a measure of convergence. An Rhat value of 1.0
    indicates perfect convergence, meaning the chains for the MCMC
    sampling are mixing well and that the estimates for each parameter
    are reliable. All Rhat values here are exactly 1.0, which suggests
    that the sampling chains converged properly.

-   `n_eff (Effective Sample Size)`: This is a measure of how many
    independent samples are available for each parameter. Larger values
    indicate better sampling. Here, most parameters have high effective
    sample sizes (ranging from \~1600 to \~3000), suggesting that the
    model has gathered a sufficiently large and diverse set of samples
    for each parameter, ensuring robust inference.

```{r}
#| label: tbl-ppd
#| tbl-cap: Model posterior predictive distribution
#| eval: true
#| echo: false
#| message: false
#| warning: false

model_summary <- summary(model)
model_summary <- as.data.frame(model_summary)

# Extract relevant fit diagnostics
fit_diagnostics <- model_summary[c("mean_PPD"),c(1, 3:6)]
fit_diagnostics$variable <- "mean_PPD"
fit_diagnostics <- fit_diagnostics[, c("variable", "mean", "sd", "10%", "50%", "90%")]

# fit_diagnostics |>
#   tt()
#   style_tt(j = 2:5, align = "r") |>   # Align columns 2 to 5 to the right
#   format_tt(digits = 1, num_fmt = "decimal")  # Format numbers with 1 decimal

# Use kableExtra for table styling
fit_diagnostics_rounded <- round(fit_diagnostics[, 2:6], 2)

fit_diagnostics_rounded |>
  kable(format = "latex")

```{r}
#| label: tbl-mcmc
#| tbl-cap: Model MCMC statisics
#| eval: true
#| echo: false
#| message: false
#| warning: false

mcmc_row_names <- c("(Intercept)", "year", "team2Bangladesh", "team2England", "team2India", "team2Ireland", "team2New Zealand", "team2South Africa", "team2Sri Lanka", "team2West Indies", "team2Zimbabwe", "toss_win")

model_summary2 <- summary(model)
model_summary2 <- as.data.frame(model_summary2)
fit_diagnostics2 <- model_summary2[mcmc_row_names,c(2, 7:8)]
fit_diagnostics2$variable <- mcmc_row_names
fit_diagnostics2 <- fit_diagnostics2[, c("variable", "mcse", "Rhat", "n_eff")]

# fit_diagnostics2 |>
#   tt() |>
#   style_tt(j = 2:4, align = "r") |>
#   format_tt(digits = 1, num_fmt = "decimal")

# Use kableExtra for table styling
fit_diagnostics_rounded2 <- round(fit_diagnostics2[, 2:4], 2)

fit_diagnostics_rounded2 |>
  kable(format = "latex")
```
